---
title: "Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(viridis)
library(ggridges)
library(lubridate)
library(stringr)
library(kableExtra)
```

```{r Data Import Function, include=FALSE, message =FALSE}
# Removed col names type for now and went with automatic column parsing

citibike_import = function(file) {
  
  df = read_csv(file, col_names = TRUE) %>%
    janitor::clean_names() %>% 
    separate(start_time, into = c("start_date", "start_time"), sep = " ") %>% 
    separate(stop_time, into = c("stop_date", "stop_time"), sep = " ") %>% 
    separate(start_date, into = c("start_year", "start_month", "start_day"), sep = "-") %>% 
    separate(stop_date, into = c("stop_year", "stop_month", "stop_day"), sep = "-") %>% 
    mutate(user_type = tolower(user_type),
           gender = as.factor(gender - 1),
           start_month = ifelse(start_month == "01", "January", "March"),
           stop_month = ifelse(start_month == "01", "January", "March"),
           trip_minutes = (trip_duration / 60),
           rider_age = (2017 - birth_year),
           start_hour = as.numeric(str_sub(start_time, 0, 2)),
           overnight = ifelse(start_hour < 06 & start_hour >= 00, 1, 0),
           morning = ifelse(start_hour < 12 & start_hour >= 06, 1, 0),
           afternoon = ifelse(start_hour < 18 & start_hour >= 12, 1, 0),
           night = ifelse(start_hour <= 24 & start_hour >= 18, 1, 0)) %>% 
    filter(user_type == "subscriber",
           trip_minutes > 5,
           trip_minutes <= 45,
           !start_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           !end_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           gender != -1,
           rider_age < 100)
  
  df
  
}

citibike_tidy = 
  bind_rows(citibike_import("./data/201701-citibike-tripdata.csv.zip"),
            citibike_import("./data/201703-citibike-tripdata.csv.zip"))

# Reading the data directly from website doesn't work - thousands of parsing failures, both with auto parsing and when specifing column parsing in function
# citibike_import("https://s3.amazonaws.com/tripdata/201701-citibike-tripdata.csv.zip"),
# citibike_import("https://s3.amazonaws.com/tripdata/201703-citibike-tripdata.csv.zip"))

```

## Exploratory Analysis

First, we restricted our analyses to users who pay for subscriptions and to trips lasting longer than 5 minutes and shorter than or equal to 45 minutes. 

### Most (and least) used starting stations

```{r Starting Station Kables}

# Top 10 starting stations in the month of January

citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(start_station_id, start_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(10, n) %>% 
  arrange(desc(n)) %>% 
  mutate(n = format(n, big.mark = ",")) %>%
  select(-start_station_id) %>% 
  rename("Starting CitiBike Station (Address)" = start_station_name,
         "Times used in January 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")

# Top 10 starting stations in the month of March

citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(start_station_id, start_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(10, n) %>% 
  arrange(desc(n)) %>%
  mutate(n = format(n, big.mark = ",")) %>%
  select(-start_station_id) %>% 
  rename("Starting CitiBike Station (Address)" = start_station_name,
         "Times used in March 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")

# Bottom 10 starting stations in the month of January

citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(start_station_id, start_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(-10, n) %>% 
  arrange(n) %>% 
  select(-start_station_id) %>% 
  rename("Starting CitiBike Station (Address)" = start_station_name,
         "Times used in January 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")

# Bottom 10 starting stations in the month of March

citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(start_station_id, start_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(-10, n) %>% 
  arrange(n) %>% 
  select(-start_station_id) %>% 
  rename("Starting CitiBike Station (Address)" = start_station_name,
         "Times used in March 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")
```


### Most (and least) used terminal stations

```{r Ending Station Kables}
# Top 10 terminal stations in the month of January

citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(end_station_id, end_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(10, n) %>% 
  arrange(desc(n)) %>% 
  mutate(n = format(n, big.mark = ",")) %>%
  select(-end_station_id) %>% 
  rename("Terminal CitiBike Station (Address)" = end_station_name,
         "Times used in January 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")

# Top 10 terminal stations in the month of March

citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(end_station_id, end_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(10, n) %>% 
  arrange(desc(n)) %>% 
  mutate(n = format(n, big.mark = ",")) %>%
  select(-end_station_id) %>% 
  rename("Terminal CitiBike Station (Address)" = end_station_name,
         "Times used in March 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")

# Bottom 10 terminal stations in the month of January

citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(end_station_id, end_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(-10, n) %>% 
  arrange(n) %>% 
  select(-end_station_id) %>% 
  rename("Terminal CitiBike Station (Address)" = end_station_name,
         "Times used in January 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")

# Bottom 10 terminal stations in the month of March

citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(end_station_id, end_station_name) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  top_n(-10, n) %>% 
  arrange(n) %>% 
  select(-end_station_id) %>% 
  rename("Terminal CitiBike Station (Address)" = end_station_name,
         "Times used in March 2017 (n)" = n) %>% 
  knitr::kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center")
```


## Linear regression

* Simple linear regression of trip duration vs. age, sex, time of day (categorical: morning rush hour, morning, evening rush hour), day of week

```{r}
# exploring a model for predicting trip duration

lm(trip_duration ~ rider_age + gender + overnight + morning + afternoon + night, 
   data = citibike_tidy) %>% 
  broom::tidy()
```

## Findings