---
title: "Data Import and Exploration"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(viridis)
library(ggridges)
library(lubridate)
library(stringr)
library(kableExtra)
```

## Introduction 

Data for this project was obtained from [the company's system data website](https://s3.amazonaws.com/tripdata/index.html), which is provided according to the [NYCBS Data Use Policy](https://www.citibikenyc.com/data-sharing-policy). We downloaded CSV zip files of Citi Bike data for the corresponding months of interest in 2017. 


***

## Data Import

Originally, we felt it was better to download directly from the data source for reproducibility purposes. However, when using read_csv with the direct URL to the zip, there were several parsing errors and the import failed. This may be due to the size of the files (zips vary from 25-70 MB). As such, we decided to download the zips of two months of interest and add them to our project's data folder and import from there.

```{r Data Import Function}
# Removed col names type for now and went with automatic column parsing

citibike_import = function(file) {
  
  df = read_csv(file, col_names = TRUE) %>%
    janitor::clean_names() %>% 
    separate(start_time, into = c("start_date", "start_time"), sep = " ") %>% 
    separate(stop_time, into = c("stop_date", "stop_time"), sep = " ") %>% 
    separate(start_date, into = c("start_year", "start_month", "start_day"), sep = "-") %>% 
    separate(stop_date, into = c("stop_year", "stop_month", "stop_day"), sep = "-") %>% 
    mutate(user_type = tolower(user_type),
           gender = as.factor(gender - 1),
           start_month = ifelse(start_month == "01", "January", "March"),
           stop_month = ifelse(start_month == "01", "January", "March"),
           trip_minutes = (trip_duration / 60),
           rider_age = (2017 - birth_year),
           start_hour = as.numeric(str_sub(start_time, 0, 2)),
           overnight = ifelse(start_hour < 06 & start_hour >= 00, 1, 0),
           morning = ifelse(start_hour < 12 & start_hour >= 06, 1, 0),
           afternoon = ifelse(start_hour < 18 & start_hour >= 12, 1, 0),
           night = ifelse(start_hour <= 24 & start_hour >= 18, 1, 0)) %>% 
    filter(user_type == "subscriber",
           trip_minutes > 5,
           trip_minutes <= 45,
           !start_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           !end_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           gender != -1,
           rider_age < 100)
  
  df
  
}

citibike_tidy = 
  bind_rows(citibike_import("./data/201701-citibike-tripdata.csv.zip"),
            citibike_import("./data/201703-citibike-tripdata.csv.zip"))

# Reading the data directly from website doesn't work - thousands of parsing failures, both with auto parsing and when specifing column parsing in function
# citibike_import("https://s3.amazonaws.com/tripdata/201701-citibike-tripdata.csv.zip"),
# citibike_import("https://s3.amazonaws.com/tripdata/201703-citibike-tripdata.csv.zip"))

```


***
