---
title: "Visualizations"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(viridis)
library(ggridges)
library(lubridate)
library(stringr)
library(kableExtra)
```

```{r Data Import Function, include=FALSE, message =FALSE}
# Removed col names type for now and went with automatic column parsing

citibike_import = function(file) {
  
  df = read_csv(file, col_names = TRUE) %>%
    janitor::clean_names() %>% 
    separate(start_time, into = c("start_date", "start_time"), sep = " ") %>% 
    separate(stop_time, into = c("stop_date", "stop_time"), sep = " ") %>% 
    separate(start_date, into = c("start_year", "start_month", "start_day"), sep = "-") %>% 
    separate(stop_date, into = c("stop_year", "stop_month", "stop_day"), sep = "-") %>% 
    mutate(user_type = tolower(user_type),
           gender = as.factor(gender - 1),
           start_month = ifelse(start_month == "01", "January", "March"),
           stop_month = ifelse(start_month == "01", "January", "March"),
           trip_minutes = (trip_duration / 60),
           rider_age = (2017 - birth_year),
           start_hour = as.numeric(str_sub(start_time, 0, 2)),
           overnight = ifelse(start_hour < 06 & start_hour >= 00, 1, 0),
           morning = ifelse(start_hour < 12 & start_hour >= 06, 1, 0),
           afternoon = ifelse(start_hour < 18 & start_hour >= 12, 1, 0),
           night = ifelse(start_hour <= 24 & start_hour >= 18, 1, 0)) %>% 
    filter(user_type == "subscriber",
           trip_minutes > 5,
           trip_minutes <= 45,
           !start_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           !end_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           gender != -1,
           rider_age < 100)
  
  df
  
}

citibike_tidy = 
  bind_rows(citibike_import("./data/201701-citibike-tripdata.csv.zip"),
            citibike_import("./data/201703-citibike-tripdata.csv.zip"))
```

## Station usage

The following plots depict the number of trips that started from and ended at each of the current Citi Bike stations, for January and March of 2017.

#### Starting station

There was little difference between station usage in the two months. The most trips started at `r citibike_tidy %>% filter(start_month == "January") %>% group_by(start_station_name) %>% summarize(n = n()) %>% top_n(1, n) %>% select(-n)`, which is located right next to Grand Central Station. There were `r citibike_tidy %>% filter(start_month == "January") %>% group_by(start_station_name) %>% summarize(n = n()) %>% top_n(1, n) %>% select(-start_station_name)` trips starting from this station in January and `r citibike_tidy %>% filter(start_month == "March") %>% group_by(start_station_name) %>% summarize(n = n()) %>% top_n(1, n) %>% select(-start_station_name)` trips in March.
  
<br/>
```{r start stations, echo=FALSE}
#plotly works nicely now

citibike_start_jan =
  citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(start_station_name, start_station_latitude, start_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", start_station_name, '\n# of trips started at this station: ', n)) %>% 
  plot_ly(x = ~start_station_longitude, y = ~start_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude")
    ) %>% 
  colorbar(title = "Jan (n)")

## map of nyc based on lat/long, overlay ggplot on top? https://cran.r-project.org/web/packages/magick/vignettes/intro.html#combining

#Jan ggplot of plotly above

#citibike_tidy %>% 
#  filter(start_month == "January") %>% 
#  group_by(start_station_id, start_station_latitude, start_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = start_station_longitude, y = start_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()

citibike_start_mar =
  citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(start_station_name, start_station_latitude, start_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", start_station_name, '\n# of trips started at this station: ', n)) %>% 
  plot_ly(x = ~start_station_longitude, y = ~start_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude")
    ) %>% 
  colorbar(title = "Mar (n)")

#March ggplot of plotly above

#citibike_tidy %>% 
#  filter(start_month == "March") %>% 
#  group_by(start_station_id, start_station_latitude, start_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = start_station_longitude, y = start_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()

subplot(
  citibike_start_jan %>% layout(showlegend = FALSE),
  citibike_start_mar %>% layout(showlegend = FALSE),
  margin = 0.05,
  shareX = TRUE, shareY = TRUE
) %>% 
layout(
  title = "Number of trips started at each station (2017)",
  annotations = list(
    list(x = 0.2 , y = 1.0, text = "January", showarrow = F, xref='paper', yref='paper'),
    list(x = 0.8 , y = 1.0, text = "March", showarrow = F, xref='paper', yref='paper'))
)
```
<br/>

#### End station

The station where the most trips ended was also `r citibike_tidy %>% filter(start_month == "January") %>% group_by(end_station_name) %>% summarize(n = n()) %>% top_n(1, n) %>% select(-n)`. In January, `r citibike_tidy %>% filter(start_month == "January") %>% group_by(end_station_name) %>% summarize(n = n()) %>% top_n(1, n) %>% select(-end_station_name)` trips ended at this station, and `r citibike_tidy %>% filter(start_month == "March") %>% group_by(end_station_name) %>% summarize(n = n()) %>% top_n(1, n) %>% select(-end_station_name)` trips ended here in March.
  
<br/>
```{r end stations, echo=FALSE}
citibike_end_jan = 
  citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(end_station_name, end_station_latitude, end_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", end_station_name, '\n# of trips started at this station: ', n)) %>% 
  plot_ly(x = ~end_station_longitude, y = ~end_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude")
    ) %>% 
  colorbar(title = "Jan (n)")

#Jan ggplot of plotly above
#citibike_tidy %>% 
#  filter(start_month == "January") %>% 
#  group_by(end_station_id, end_station_latitude, end_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = end_station_longitude, y = end_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()

citibike_end_mar = 
citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(end_station_name, end_station_latitude, end_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", end_station_name, '\n# of trips started at this station: ', n)) %>% 
  plot_ly(x = ~end_station_longitude, y = ~end_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label) %>% 
  layout(
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude")
    ) %>% 
  colorbar(title = "Mar (n)")

#March ggplot of plotly above
#citibike_tidy %>% 
#  filter(start_month == "March") %>% 
#  group_by(end_station_id, end_station_latitude, end_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = end_station_longitude, y = end_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()

subplot(
  citibike_end_jan %>% layout(showlegend = FALSE),
  citibike_end_mar %>% layout(showlegend = FALSE),
  margin = 0.05,
  shareX = TRUE, shareY = TRUE
) %>% 
layout(
  title = "Number of trips ended at each station (2017)",
  annotations = list(
    list(x = 0.2 , y = 1.0, text = "January", showarrow = F, xref='paper', yref='paper'),
    list(x = 0.8 , y = 1.0, text = "March", showarrow = F, xref='paper', yref='paper'))
)
```

<br/>

***

<br/>
  
## Rider characteristics

#### Gender and age

In the months analyzed, there were significantly more rides by males compared to females. There were a total of `r format(nrow(subset(citibike_tidy, gender == 0)), big.mark = ",")` rides taken by male riders and `r format(nrow(subset(citibike_tidy, gender == 1)), big.mark = ",")` rides taken by female riders.

The average age of a male rider was `r round(mean(citibike_tidy$rider_age[which(citibike_tidy$gender==0)]), digits = 2)` years, and ``r round(mean(citibike_tidy$rider_age[which(citibike_tidy$gender==1)]), digits = 2)` years for female riders. The oldest male rider was `r max(citibike_tidy$rider_age[which(citibike_tidy$gender==0)])` years old, and the oldest female rider was `r max(citibike_tidy$rider_age[which(citibike_tidy$gender==1)])` years old.
  
```{r riders age sex, echo=FALSE}
plot_ly(alpha = 0.6) %>%
  add_histogram(x = ~citibike_tidy$rider_age[which(citibike_tidy$gender == 0)], name = "Male") %>%
  add_histogram(x = ~citibike_tidy$rider_age[which(citibike_tidy$gender == 1)], name = "Female") %>%
  layout(barmode = "overlay",
         title = "Age of riders, by gender",
         xaxis = list(title = "Rider age (years)"),
         yaxis = list(title = "# riders (n)")
    )
```

<br/>

***

<br/>

## Trip characteristics

#### Length of ride against day of month

```{r trip_length_plot, echo=FALSE}
citibike_tidy %>% 
  mutate(start_day = as.numeric(start_day)) %>% 
  group_by(start_month, start_day) %>% 
  mutate(mean_trip_minutes = mean(trip_minutes)) %>% 
  ggplot(aes(x = start_day, y = mean_trip_minutes, color = start_month)) +
  geom_point(alpha = 0.7) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_x_continuous(breaks = seq(1, 31, 2)) + 
  scale_y_continuous(breaks = seq(11.25, 16, 0.25)) +
  labs(x = "Day of the month",
       y = "Mean length of trip (minutes)",
       title = "Mean length of rides over the course of the month") + 
  scale_color_hue(name = "Month") +
  facet_grid(~start_month)
```