---
title: "Visualizations"
html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(viridis)
library(ggridges)
library(lubridate)
library(stringr)
library(kableExtra)
```

```{r Data Import Function, include=FALSE, message=FALSE}
# Removed col names type for now and went with automatic column parsing

citibike_import = function(file) {
  
  df = read_csv(file, col_names = TRUE) %>%
    janitor::clean_names() %>% 
    separate(start_time, into = c("start_date", "start_time"), sep = " ") %>% 
    separate(stop_time, into = c("stop_date", "stop_time"), sep = " ") %>% 
    separate(start_date, into = c("start_year", "start_month", "start_day"), sep = "-") %>% 
    separate(stop_date, into = c("stop_year", "stop_month", "stop_day"), sep = "-") %>% 
    mutate(user_type = tolower(user_type),
           gender = as.factor(gender - 1),
           start_month = ifelse(start_month == "01", "January", "March"),
           stop_month = ifelse(start_month == "01", "January", "March"),
           trip_minutes = (trip_duration / 60),
           rider_age = (2017 - birth_year),
           start_hour = as.numeric(str_sub(start_time, 0, 2)),
           overnight = ifelse(start_hour < 06 & start_hour >= 00, 1, 0),
           morning = ifelse(start_hour < 12 & start_hour >= 06, 1, 0),
           afternoon = ifelse(start_hour < 18 & start_hour >= 12, 1, 0),
           night = ifelse(start_hour <= 24 & start_hour >= 18, 1, 0)) %>% 
    filter(user_type == "subscriber",
           trip_minutes > 5,
           trip_minutes <= 45,
           !start_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           !end_station_name %in% c("Bressler", "Kiosk in a box Motivate", "NYCBS Depot - FAR", "NYCBS Depot - SSP", "NYCBS Depot - STY", "NYCBS Depot - DYR", "NYCBS Depot BAL - DYR", "SSP Tech Workshop", "Penn Station Valet", "Penn Station Valet - Valet Scan"),
           gender != -1,
           rider_age < 100)
  
  df
  
}

citibike_tidy = 
  bind_rows(citibike_import("./data/201701-citibike-tripdata.csv.zip"),
            citibike_import("./data/201703-citibike-tripdata.csv.zip"))

# Reading the data directly from website doesn't work - thousands of parsing failures, both with auto parsing and when specifing column parsing in function
# citibike_import("https://s3.amazonaws.com/tripdata/201701-citibike-tripdata.csv.zip"),
# citibike_import("https://s3.amazonaws.com/tripdata/201703-citibike-tripdata.csv.zip"))

```

## Visualizations

### Station usage

The following two plots depict the number of trips that started from each of the current Citi Bike stations, for January and March of 2017.

```{r jan2017 start stations}

#plotly works nicely now

citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(start_station_name, start_station_latitude, start_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", start_station_name, '\n# of trips started at this station: ', n)) %>% 
  plot_ly(x = ~start_station_longitude, y = ~start_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label) %>% 
  layout(
    title = "Starting station usage, January 2017",
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude")
    )

## map of nyc based on lat/long, overlay ggplot on top? https://cran.r-project.org/web/packages/magick/vignettes/intro.html#combining

#Jan ggplot of plotly above

#citibike_tidy %>% 
#  filter(start_month == "January") %>% 
#  group_by(start_station_id, start_station_latitude, start_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = start_station_longitude, y = start_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()
```

In January 2017, 


``` {r}
citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(start_station_name, start_station_latitude, start_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", start_station_name, '\n# of trips started at this station: ', n)) %>% 
  plot_ly(x = ~start_station_longitude, y = ~start_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label) %>% 
  layout(
    title = "Starting station usage, March 2017",
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude")
    )

#March ggplot of plotly above

#citibike_tidy %>% 
#  filter(start_month == "March") %>% 
#  group_by(start_station_id, start_station_latitude, start_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = start_station_longitude, y = start_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()
```

```{r}
citibike_tidy %>% 
  filter(start_month == "January") %>% 
  group_by(end_station_name, end_station_latitude, end_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", end_station_name, '\n# of trips ended at this station: ', n)) %>% 
  plot_ly(x = ~end_station_longitude, y = ~end_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label)

#Jan ggplot of plotly above
#citibike_tidy %>% 
#  filter(start_month == "January") %>% 
#  group_by(end_station_id, end_station_latitude, end_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = end_station_longitude, y = end_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()

citibike_tidy %>% 
  filter(start_month == "March") %>% 
  group_by(end_station_name, end_station_latitude, end_station_longitude) %>% 
  summarize(n = n()) %>% 
  mutate(text_label = str_c("Station name: ", end_station_name, '\n# of trips ended at this station: ', n)) %>% 
  plot_ly(x = ~end_station_longitude, y = ~end_station_latitude, type = "scatter", mode = "markers",
          alpha = 0.5,
          color = ~n,
          text = ~text_label)

#March ggplot of plotly above
#citibike_tidy %>% 
#  filter(start_month == "March") %>% 
#  group_by(end_station_id, end_station_latitude, end_station_longitude) %>% 
#  summarize(n = n()) %>% 
#  ggplot(aes(x = end_station_longitude, y = end_station_latitude, color = n)) +
#  geom_point(alpha = 0.3) + 
#  scale_color_viridis() +
#  coord_cartesian() +
#  theme_classic()
```


### Trips by gender (1 = Male, 2 = Female)

```{r gender}
# Histogram of when people start their rides, by gender!
citibike_tidy %>% 
  group_by(gender) %>% 
  mutate(start_hour = str_sub(start_time, 0, 2),
         start_hour = as.numeric(start_hour)) %>% 
  ggplot(aes(x = start_hour)) +
  geom_histogram() +
  labs(x = "Start hour",
       y = "Frequency") +
  facet_grid(~ gender)
```

### Age ranges and average time spent on ride

```{r age_length_plot}
# who are the people over 100; also, this is kind of ug the way i have done it; whose trips are more than 500 hours??????? how is this calculated??
# Will definitely need to look into how some of these indicators are measured/calculated... This plot is pretty useless as is.
citibike_tidy %>% 
  mutate(trip_hour = trip_minutes / 60) %>% 
  ggplot(aes(x = rider_age, y = trip_hour)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(~ gender)
```

### Most/least popular days of the week of travel 

* Simple linear regression of trip duration vs. age, sex, time of day (categorical: morning rush hour, morning, evening rush hour), day of week

```{r}
# exploring a model for predicting trip duration

lm(trip_duration ~ rider_age + gender + overnight + morning + afternoon + night, 
   data = citibike_tidy) %>% 
  broom::tidy()
```

* Length of ride against day of month, by month

```{r trip_length_plot}
# Histogram of how long people ride for, by month!
# can we change the month labels??

  citibike_tidy %>% 
  mutate(start_day = as.numeric(start_day)) %>% 
  group_by(start_month, start_day) %>% 
    mutate(mean_trip_hour = mean(trip_minutes / 60)) %>% 
    ggplot(aes(x = start_day, y = mean_trip_hour, color = start_month)) +
    geom_point(alpha = 0.5) +
    geom_smooth(se = FALSE) +
    labs(x = "Day of the month",
         y = "Mean length of trip (hours)") + 
    scale_color_hue(name = "Month") +
    facet_grid(~ start_month)
```
